The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-11-30 article_final/main.tex
Expanded appendix

The extremal members of this family are \emph{F-Neon} (with $k = 1$) and \emph{F-Muon} (with $k = \min\{m, n\}$). Additional information and visualizations for the $\normfstar{\cdot} = \norm{\cdot}_{\mathrm{F-KF-1}}$ norm appear in the appendix (see \equationref{eq:normfstardual}, \figureref{fig:fstardual}).

@@

Notably, even the pathological case $\alpha > 1$—which corresponds to an LMO constraint region that is not a norm ball—achieves accuracy nearly identical to vanilla Muon.

@@

@@

\section{Norms $\normfstar{\cdot}^\dagger$ and $\normftwo{\cdot}^\dagger$}

@@

Unfortunately, $\normfstar{\cdot}^\dagger$ and $\normftwo{\cdot}^\dagger$ do not have a closed-form expression.

@@

\section{More data for Linear Least Squares}
\label{sec:lls_plots_section}

\begin{table}[htbp]
\floatconts
  {tbl:lls_lrs_app}
  {\caption{Tuning lrs and momentum coefficients}}
  {%
    \begin{tabular}{|c|c|c|c|}
      \arrayrulecolor{black}\hline
      Algorithm & \texttt{lr}  & \texttt{momentum}  & Iterations to 0.001 loss \\
      \arrayrulecolor{black}\hline
      Muon & 0.007 & 0.5 & 1060 \\
      \hdashline
      NSGD & 0.08 & 0.95 & 1020 \\
      F-Muon & 0.015 & 0.7 & 910 \\
      \hdashline
      SignSGD & 0.016 $\times$ 0.01 & 0.95 & 2650 \\
      S-Muon & 0.011 & 0.9 & 890 \\
      \arrayrulecolor{black}\hline
    \end{tabular}%
  }
\end{table}

\begin{figure}[htbp]
\floatconts
  {fig:app_lls}
  {\caption{More images for Linear least squares problem for a 500x500 matrix}}
  {%
    \subfigure[The spectral norm of the gradient]{\label{fig:lls_op_grad_norm}%
      \includegraphics[width=0.5\linewidth]{figs/simple_lls/gradient_spectral_norm_vs_iteration_long_run.pdf}}%
    \vspace{1em}
    \subfigure[The nuclear norm of the gradient]{\label{fig:lls_nuclear_grad_norm}%
      \includegraphics[width=0.5\linewidth]{figs/simple_lls/gradient_nuclear_norm_vs_iteration_long_run.pdf}}%
    \vspace{1em}
    \subfigure[The loss over time]{\label{fig:lls_loss_time}%
      \includegraphics[width=0.5\linewidth]{figs/simple_lls/loss_vs_time_long_run.pdf}}
  }
\end{figure}

@@

We compare Muon, F-Muon, S-Muon, NSGD, SignSGD, Neon, F-Neon, Fanion-5, and F-Fanion-5 on CIFAR-airbench. NSGD and SignSGD are not heavily tuned. The hyperparameters of Neon, F-Neon, Fanion-5, and F-Fanion-5 are taken from Muon and F-Muon (see \tableref{tbl:cifar_lrs}). In the table, val\_accuracy is for the the weighted variant. However, to preserve the algorithms, we do not normalize the weights of the network each iteration.

We log train and validation accuracy, and nuclear, gradient, and spectral norms of the gradients of conv1.weight and conv2.weight for all the layers. The gradient norms decrease only by the order of ten maximum during training, while train accuracy reaches full 100\% for Muon, S-Muon, and F-Muon.

@@

\paragraph{CIFAR airbench} We used NVIDIA RTX A4000.

\paragraph{NanoGPT pretrain} We used the standard setting of 8 $\times$ H100 as documented in \citep{modded_nanogpt_2024}.

>>>>

The extremal members of this family are \emph{F-Neon} (with $k = 1$) and \emph{F-Muon} (with $k = \min\{m, n\}$). Additional information and visualizations for the F-Muon generating $\normfstar{\cdot}^\dagger = \norm{\cdot}_{\mathrm{F-KF-1}}^\dagger$ and the F-Neon generating $\normftwo{\cdot}^\dagger = \norm{\cdot}_{\mathrm{F-KF-\min\{m, n\}}}^\dagger$ norms appear in the appendix (see \equationref{eq:normfstardual}, \equationref{eq:normftwodual}, \figureref{fig:fduals}).

@@

Notably, even the pathological case $\alpha > 1$, which corresponds to an LMO constraint region that is not a norm ball, achieves accuracy nearly identical to vanilla Muon.

@@

For gradient norm profiling and results for training with Fanions and F-Fanions, please see \sectionref{sec:cifar_underhood}.

@@

\section[Norms for F-Muon and F-Neon]%
        {Norms $\normfstar{\cdot}^\dagger$ and $\normftwo{\cdot}^\dagger$}
\label{sec:f2_fstar_norms}

@@

Unfortunately, $\normfstar{\cdot}^\dagger$ and $\normftwo{\cdot}^\dagger$ do not have closed-form expressions and cannot be found as easy as their duals.

@@

\section{More details for Linear Least Squares}
\label{sec:lls_plots_section}
Tuned \texttt{lr} and \texttt{momentum} coefficients are in the \tableref{tbl:lls_lrs_app}. Spectral, nuclear norm of the gradients over iterations and loss over time are in the \figureref{fig:app_lls}. Very low speed of Fanions and F-Fanions for large $k$ is probably caused by ineffective implementation of TRLan by \cite{cupy_svds_ref}, since $\mathcal{O}(mn^2 + n^2k + nk^2)$ complexity is actually $\mathcal{O}(mn^2)$.

\begin{table}[htbp]
\floatconts
  {tbl:lls_lrs_app}
  {\caption{Tuning \texttt{lr} and \texttt{momentum} coefficients}}
  {%
    \begin{tabular}{|c|c|c|c|}
      \arrayrulecolor{black}\hline
      Algorithm & \texttt{lr}  & \texttt{momentum}  & Iterations to 0.001 loss \\
      \arrayrulecolor{black}\hline
      Muon & 0.007 & 0.5 & 1060 \\
      \hdashline
      NSGD & 0.08 & 0.95 & 1020 \\
      F-Muon & 0.015 & 0.7 & 910 \\
      \hdashline
      SignSGD & 0.016 $\times$ 0.01 & 0.95 & 2650 \\
      S-Muon & 0.011 & 0.9 & 890 \\
      \arrayrulecolor{black}\hline
    \end{tabular}%
  }
\end{table}

\begin{figure}[htbp]
\floatconts
  {fig:app_lls}
  {\caption{More plots for linear least squares problem for a 500x500 matrix}}
  {%
    \subfigure[The spectral norm of the gradient]{\label{fig:lls_op_grad_norm}%
      \includegraphics[width=0.5\linewidth]{figs/simple_lls/gradient_spectral_norm_vs_iteration_long_run.pdf}}%
    \vspace{1em}
    \subfigure[The nuclear norm of the gradient]{\label{fig:lls_nuclear_grad_norm}%
      \includegraphics[width=0.5\linewidth]{figs/simple_lls/gradient_nuclear_norm_vs_iteration_long_run.pdf}}%
    \vspace{1em}
    \subfigure[The loss over time]{\label{fig:lls_loss_time}%
      \includegraphics[width=0.5\linewidth]{figs/simple_lls/loss_vs_time_long_run.pdf}}
  }
\end{figure}

@@

We compare Muon, F-Muon, S-Muon, NSGD, SignSGD, Neon, F-Neon, Fanion-5, and F-Fanion-5 on CIFAR-airbench. NSGD and SignSGD are not heavily tuned. The hyperparameters of Neon, F-Neon, Fanion-5, and F-Fanion-5 are taken from Muon and F-Muon (see \tableref{tbl:cifar_lrs}). In the table, val\_accuracy after 8 epochs is for the the airbench variant with weight normalization at each step. However, in order to be faithful to the conditions of the convergence theorems \citep{kovalev2025understanding,riabinin2025gluon}, we do not normalize the weights of the network when logging gradient norms.

We log train and validation accuracies (\figureref{fig:cifar_accs}), and nuclear, gradient, and spectral norms of the gradients of conv1.weight and conv2.weight for all the layers (see \figureref{fig:norms_conv1_vs_conv2} for norms in the layer 2, and \figureref{fig:total_norms_cifar} for total norms of the weights). During training, the gradient norms decrease only by the order of ten maximum, while train accuracy reaches full 100\% for Muon, S-Muon, and F-Muon. This means that the bounds of the $\max\{\frac{A}{\epsilon}, \frac{B}{\epsilon^2}, \frac{C}{\epsilon^3}, \frac{D}{\epsilon^4}\}$ type (like Corollary 2 from \citet{kovalev2025understanding}) probably should not be reduced to $\frac{D}{\epsilon^4}$ when selecting the optimal algorithm for a practical problem.

@@

\paragraph{Linear Least Squares and CIFAR airbench} Experiments were conducted on a single NVIDIA RTX A4000. The code was run with Python 3.10.13 and Pytorch~2.6.0+cu126.

\paragraph{NanoGPT and GPT-2 Medium Pretrain} We used the standard setting of 8 $\times$ NVIDIA H100 as documented in \citep{modded_nanogpt_2024}. The code was run with Python 3.10.18, Pytorch~2.10.0.dev20251124+cu126 on a computer with Xeon® Platinum 8462Y+ processor and 130 GB of RAM.
